---
title: "Using knitrdrawio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using knitrdrawio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # We need to produce figures in `vignettes` for pkgdown.
  fig.path = "."
)

library(knitrdrawio)
```

# Overview

The **knitrdrawio** package adds a `drawio` engine to [knitr], which can be
used to automatically export [draw.io] diagrams into images from code chunks
in RMarkdown (Rmd) documents.

You can thus include both your diagrams' source files and the process to
produce the diagrams in your computational documents.

# Installation

**knitrdrawio** is only available on GitHub, you must therefore use `remotes`,
or `devtools` to install it.

Using `remotes`:
```r
install.packages("remotes")
remotes::install_github("rchaput/knitdrawio")
```

**knitrdrawio** also needs you to have the [draw.io] desktop app installed on
your system, which you can download directly from their
[GitHub releases](https://github.com/jgraph/drawio-desktop/releases).

# Usage

## Loading the package

**knitrdrawio** must be loaded in your Rmd document in order for [knitr] to
recognize the new engine. Simply add a code chunk somewhere at the beginning
of your document:

    ```{r}`r ''`
    library(knitrdrawio)
    ```

If you do not include the `library(knitrdrawio)` line before you use a
`drawio` chunk, knitr will complain and produce the following warning:

> Warning in get_engine(options\$engine) :
>   Unknown language engine 'drawio' (must be registered via knit_engines\$set()).

In addition, the chunk (and thus your diagram) will not be rendered.

## Rendering a diagram

To produce a diagram within a Rmd document, put a new chunk using the
`drawio` engine, i.e., setting `drawio` as the first word inside the
brackets:

    ```{drawio diag1, src = "/path/to/diagram.drawio"}`r ''`
    ```

In this example, `diag1` is the label, or name, of your diagram. This name
is used in the [knitr] progress report in your console, log messages,
as well as in the filename of the produced diagram.

If you do not specify it, [knitr] will generate one label for you, such as
`unnamed-chunk-1`, which works fine, but makes harder to identify which file
corresponds to which diagram.

The `src` chunk option is relative to **knitrdrawio** and is *mandatory*:
it specifies the path to your source diagram, usually a `.drawio` file.

With the previous code chunk, and assuming that the source file exists, and
is a correct diagram, when the document is rendered, **knitrdrawio** will
automatically call draw.io to render the diagram. The resulting image
will be included in the document, as the following example shows:

```{drawio example-workflow, src = "workflow.drawio"}
```

# Adding options

**knitrdrawio** supports many options to control its behaviour.

The export process itself can be controlled by options, which are quite
similar to those offered by draw.io when exporting a diagram on the desktop
app.
Such options include the ability to *crop* a document, to choose the
*page index* to export, to add a *border*, etc.

The chunk and how it will be included in the document can be controlled by
several [knitr options][knitr-options].
Note that **knitrdrawio** was not tested against all of these options, and
some of them might not work.

## Drawio-specific options

We list here the chunk options that can be used in **knitrdrawio**, their
draw.io equivalent, and default values if left unspecified.

For example, the `page.range` option can be used as such:

    ```{drawio diag-with-page-range, src = "/path/to/diagram.drawio", page.range = "2..4"}`r ''`
    ```

```{r drawio-options-table}
kable(data.frame(
  option = c("src", "format", "crop", "transparent", "border", "page.index", "page.range"),
  equivalent = c("`input file`", "-f --format", "--crop", "-t --transparent", "-b --border", "-p --page-index", "-g --page-range"),
  desc = c("Path to the source diagram to export, usually a .drawio, .xml, or .vsdx file.", "Export format to use. Possible values are: pdf, png, jpg, svg, vsdx, xml.", "Whether to crop resulting image to diagram size.", "Set transparent background for PNG.", "Set border width around diagram. Non-zero values show the border, 0 to disable border.", "Select a specific page to export.", "Similar to page.index, but selecting a range instead. For PDF only. Usage: from..to"),
  default = c("(Mandatory argument)", "Depends on document output format: latex => pdf, html => svg, other => png", "TRUE", "FALSE", "0", "If output format is an image => 1st page, otherwise => all pages are exported.", "NULL")
  ),
  col.names = c("Chunk option", "Drawio equivalent", "Description", "Default value"))
```

For more details, please see the draw.io command line help, with `drawio --help`.

## Knitr options

**knitrdrawio** uses knitr chunks, and thus rely on the chunk options.
We only list a few options of interest below.

### Labelling the diagram chunk

The `label` chunk option controls the name that will be given to the resulting
image. By default, [knitr] will set a label if left unspecified, such as
`unnamed-chunk-1`, but this makes harder to identify the file that corresponds
to a diagram. In addition, if the chunk rendering fails, it is harder to
identify from the log which of your diagrams has an error.

Note that [knitr] allows a special syntax for this option, as the `label=`
part can be skipped.
Thus, instead of writing `{drawio label=diag1}`, we use in our examples the
simpler `{drawio diag1}`.

### Naming and referencing the diagram figure

As the diagram is included as an image in the Rmd document, it can be named
and referenced just like other images, by setting the "figure caption" option
`fig.cap`.

Depending on your document output format, the caption will appear somewhere
near the figure, e.g., below, above, or in the margin.

See for example the following chunk and its result:

    ```{drawio test-fig-cap, src = "workflow.drawio", fig.cap = "An example workflow."}`r ''`
    ```

```{drawio test-fig-cap, src = "workflow.drawio", fig.cap = "An example workflow.", eval = TRUE}
```

*Note*: at the time of writing, it is not possible to split the chunk option
line.
This means that long `fig.cap` can be hard to read.
Fortunately, [knitr] proposes an alternative syntax, based on YAML, to specify
chunk options in the chunk itself.
See the following chunk example:

    ```{drawio test-fig-cap-long, src = "workflow.drawio"}`r ''`
    #| fig.cap: >
      A very, very long sentence that shows how we can use YAML syntax to write
      long figure captions directly in the chunk itself.
    ```

### Setting the path to drawio

In most cases, **knitrdrawio** should be able to find your installation of
draw.io.
However, if for some reason the detection fails, **knitrdrawio** will not
work, and print an error message, prompting you to manually set the path to
draw.io in the chunk options, by setting the `engine.path` option.

You can set it directly in each of your chunks, but this is rather cumbersome ;
we recommend setting it for all chunks, using the `knitr::opts_chunks$set`
function.
To do so, place the following code in a chunk at the beginning of your
document (it may be the same chunk used to load **knitrdrawio**):

```r
knitr::opts_chunks$set(engine.path = list(
  drawio = "/path/to/your/draw.io"
))
```

### Change the figure folder

By default, **knitrdrawio** will produce diagrams in the folder that knitr
uses for figures (as diagrams are a special kind of figure).
This folder can be controlled by the `fig.path` option.

For example, if you want to create the diagram in the `figures/diagrams`
folder, you may use:

    ```{drawio diag2, src = "/path/to/diagram.drawio", fig.path = "figures/diagrams"}`r ''`
    ```

Non-existing directories are automatically created if necessary.

### Cache the result

Exporting a diagram can be a long operation, especially as the size of your
diagram grows.
To avoid losing too much time, **knitrdrawio** may use the caching mechanism
of knitr, if you set the `cache` chunk option to `TRUE`.

In this case, the produced diagram will be kept, and associated with a hash
key, which determines whether the diagram should be rebuilt or not.

The hash key is invalidated when:
- the source diagram changes ;
- the chunk itself is changed.

### Evaluate and/or include the diagram

As all knitr chunks, you may choose whether to evaluate and/or include a
`drawio` chunk, by setting respectively the `eval` and `include` options.

When `eval` is set to `FALSE`, the chunk will not be executed, and the diagram
not rendered.
This may be useful if you know there is an error in your diagram, and you do
not wish to crash your Rmd document.

When `include` is set to `FALSE`, the diagram is rendered, and the
corresponding file is created in the figures directory, but the figure is not
included in the Rmd document.
This may be useful if you want to manually insert it later, or simply to
produce the diagram to have it available on disk, but not actually use it.

### Setting manually the drawio arguments

You can manually pass additional command-line arguments to the `draw.io`
binary, by setting the standard `engine.opts` option.

Although most of the drawio arguments should be handled by **knitrdrawio**'s
options, this is useful if one of your cases is not covered.

Note that you can set this option globally through the `knitr::chunks_opt$set`
function, exactly as for the `engine.path` option.


[knitr]: https://yihui.org/knitr/
[knitr-options]: https://yihui.org/knitr/options/
[draw.io]: https://diagrams.net/
